---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![R build status](https://github.com/joeroe/era/workflows/R-CMD-check/badge.svg)](https://github.com/joeroe/era/actions)
[![Codecov test coverage](https://codecov.io/gh/joeroe/era/branch/master/graph/badge.svg)](https://codecov.io/gh/joeroe/era?branch=master)
<!-- badges: end -->

**era** is an R package that provides a consistent vector representation of years with an associated calendar era or time scale.
It includes built-in definitions of many contemporary and historic calendars;
time scales commonly used in archaeology, astronomy, geology, and other palaeosciences (e.g. Before Present, SI-prefixed *annus*);
and support for arbitrary user-defined eras. 
Functions for converting between eras and for type-stable arithmetic with years are also provided.

## Installation

You can install the development version of era from GitHub with the [remotes](https://remotes.r-lib.org/) package.

```{r install, eval=FALSE}
# install.packages("remotes")
remotes::install_github("joeroe/era")
```

## Basic usage

`yr()` defines the era associated with a vector of years:

```{r eg-yr}
library(era)
yr(10010:10001, "cal BP")
```

`yr` is a vector class is based on [vctrs](https://vctrs.r-lib.org/).
This means it behaves in a consistent, type-stable way across base R and other packages, and fits neatly into tibbles and data frames.
Many common calendar systems and time scales are built into the package (see `?eras()`) and can be referenced by simply passing their abbreviated label to `yr()`.
Other eras can be defined using the `era()` function directly.

Use `yr_transform()` to convert between eras:

```{r eg-transform}
yr(10010:10001, "cal BP") %>% 
  yr_transform(era("BCE"))
```

For further usage, see the [package introduction](https://era.joeroe.io/articles/era.html) (`vignette("era")`).
