---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# era <a href='https://era.joeroe.io'><img src='man/figures/logo.svg' align="right" style="float: right; height: 139px" /></a>

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![CRAN status](https://www.r-pkg.org/badges/version/era)](https://CRAN.R-project.org/package=era)
[![CRAN checks](https://badges.cranchecks.info/worst/era.svg)](https://cran.r-project.org/web/checks/check_results_era.html)
[![R-CMD-check](https://github.com/joeroe/era/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/joeroe/era/actions/workflows/R-CMD-check.yaml)
[![Test coverage](https://codecov.io/gh/joeroe/era/graph/badge.svg)](https://app.codecov.io/gh/joeroe/era)
<!-- badges: end -->

**era** is an R package that provides a consistent representation of year-based time scales as a numeric vector with an associated *era* – the [yr class](https://era.joeroe.io/reference/yr.html).

It includes built-in [era definitions](https://era.joeroe.io/reference/eras.html) for many year numbering systems used in contemporary and historic calendars  (e.g. Common Era, Islamic 'Hijri' years); year-based time scales used in archaeology, astronomy, geology, and other palaeosciences (e.g.  Before Present, SI-prefixed *annus*); and support for [arbitrary user-defined eras](https://era.joeroe.io/reference/era.html). Years can converted from any one era to another using the generalised  transformation function [yr_transform()](https://era.joeroe.io/reference/yr_transform.html).

era's classes are based on [vctrs](https://vctrs.r-lib.org/), and come with methods for robust casting and coercion between years and other numeric types, type-stable arithmetic with years, and pretty-printing in tables.

## Installation

You can install the released version of era [from CRAN](https://cran.r-project.org/package=era) with:

```{r install, eval=FALSE}
install.packages("era")
```

Or the development version from [GitHub](https://github.com/joeroe/era) using the [remotes](https://remotes.r-lib.org/) package:

```{r install-dev, eval=FALSE}
# install.packages("remotes")
remotes::install_github("joeroe/era")
```

## Basic usage

`yr()` defines the era associated with a vector of years:

```{r eg-yr}
library(era)
x <- yr(c(9000, 8000, 7000), "cal BP")
x
```


Use `yr_transform()` to convert between eras:

```{r eg-transform}
yr_transform(x, "BCE")
```

Many common calendar systems and time scales are predefined (see `?eras()`) and can be referenced by their abbreviated labels.
Other eras can be defined using the `era()` function directly.

For further usage, see the [package introduction](https://era.joeroe.io/articles/era.html) (`vignette("era")`).
